---
title: "Analysis"
author: "Courtney Fitzgerald"
date: "2024-10-03"
output: html_document
---
#Introduction
#### Drug overdoses in the US have been steadily increasing over the years. However the distribution of these deaths are not uniform for everyone. Certain groups, whether they be based around age, race, or gender, can be more prone than others to have a drug over dose. To better understand which groups are most at risk and for what drugs they are most likely to be impacted by we perform multiple regressions on data taken from the National Center for Health Staitstics and specifically their National Vital Statistics System. These Data were gathered in 2019. These specific data are labelled as Table 008 on the Center for Disease Control's website. https://www.cdc.gov/nchs/hus/data-finder.htm?year=2019

#The Data
####First we must load the data in the proper format and get a grasp of the distribution for variables of interest. We must also load in the appropriate packages to clean and analyze the data like, tidyverse and dpylr. 
```{r setup, include=FALSE}
drug_od <- read_csv("drug_data.csv")
install.packages("tidyverse")
library(tidyverse)
install.packages("ggplot")
```


##Examining the Data
####In order to evaluate how to best analyze the data we must familiarize ourselves with the variables and how these variables are oriented. This data set classifies drug overdose by type, and also takes into account estimations for the rate of overdose per 100,000 residents  for group listed. 
```{r}

table(drug_od$PANEL)
table(drug_od$PANEL_NUM)
##This gives us all of the drug overdoses caused by opioids from 1999 - 2018 and groups them by gender/racial identity. 

opioids1 <- drug_od %>% 
  filter(PANEL_NUM ==1) %>%
  filter(STUB_NAME_NUM == 4) %>%
  filter(YEAR == 2003) %>%
  mutate(Race_Gender = STUB_NAME_NUM) 


##use mutate to separate race and gender? 
##Stub_name corresponds to sex or racial grouping, stub_label specifies the group (i.e. female/male, white/black/etc)
###stub_name_num, 0 = total, 1 = age range, 2 = sex, 3 = sex/age range, 4 = sex+race, 5 = sex/ethnicity
###stub_label_num has tenths decimals that correspond to specific pairings (i.e. 4.5 = female and white)
  
```

```{r}
```

```{r}

```


```{r}
install.packages("tidyverse")
install.packages("renv")
library(tidyverse)
suicide_data <- read_csv("suicide_data.csv")
suicide1 <- read_csv("data/pc.csv")
drug_od <- read_csv("drug_data.csv")
rm(drug_data)
##use ".." to get to project level data 
##put "data/" before csv file name if data is in "data" folder 
renv::snapshot()
renv::restore()
renv::hydrate()

##GH:hosting"remotes"
##clone:local copy of remote
##remote is "origin"
##fork: copy repo on GH
##commit: creating a snapshot of changes
##push: update remote repository with local changed
##can't always push after commit, need to pull before pushing to update repo
##pull: update local repo with commits pushed by others 
```

##

You can also embed plots, for example:

```{r}
table(drug_od$PANEL)
table(drug_od$PANEL_NUM)
```

```{r}
```

```{r}
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
