---
title: "EDA Code Notebook"
author: "Courtney Fitzgerald and Martinez Jean Claude"
date: "2024-10-22"
output: html_document
---
## Introduction

#### Background

Opioids are a serious problem in the United States. Since the late 1990's the **opioid crisis has steadily worsened**. Many attribute the origins of this epidemic to the "Pain as the 5th Vital Sign" campaign and the FDA's approval of the infamous Oxycontin drug. While the opioid crisis is considered a public health emergency, it is not uniformly harmful for everyone. Like many epidemics, **certain demographic groups are more at risk than others**. 

#### Purpose 

Understanding **how and why certain groups are most at risk of dying from opioids** is the first step to reducing the magnitude of this epidemic. This information will **enable targeted interventions**, whether they be at the community or policy level. 

#### Objectives

This project **will examine some of the demographic factors, like age, race/ethnicity, and gender, that contribute to and influence the harrowing opioid epidemic in the United States**. Through this demographic analysis, we will examine what cultural and societal factors lead to variations between certain groups. 

#### Research Question 

Specifically, we want to answer; *How do race, ethnicity, and gender contribute to variations in opioid-related mortalities?* 


## The Data 


#### Source

The data employed in this study is taken from the National Center for Health Statistics and specifically their National Vital Statistics System. 

The data were gathered in 2019 and is based off of information taken from death certificates.The official information provided about the data set on the Center for Disease Control's website is, "[Drug overdose death rates, by drug type, sex, age, race, and Hispanic origin: United States, selected years 1999-2018](https://data.cdc.gov/NCHS/Drug-overdose-death-rates-by-drug-type-sex-age-rac/95ax-ymtc/about_data)". 

We use this dataset because it provides the demographic variables we need to explore between group comparisons, and it also provides an estimated rate of deaths per 100,000 for a given group. As such, the data's appropriate combination of numerical and categorical variables will allow us to perform the appropriate statistical tests given our independent variables of race, ethnicity, and gender, and our single dependent variable of estimated death rate.

**Additional information:**
The data in CSV file format can be found [here](https://www.cdc.gov/nchs/hus/contents2019.htm#Table-008).

**Variables** | **Description**  
--- | ---  
Age | Ages of the observations but not specified in the dataset 
Sex | Male or Female   
Ethnicity | Hispanic non-specified race or Non-Hispanic    
Race_Ethnicity | NH-Black or NH-White or NH-Native American or NH-AAPI or Hispanic non-specified race     
Drug Type | Natural/Semisynthetic Opioid or Methadone or Heroin or Synthetic Opioid   
Year | 1999 - 2018   
Estimate | Rate of Overdose Deaths per 100,000     

#### Data Limitations

While these data do include information on race, gender, and sex, they are not aggregated to our liking:

1. **When Hispanic ethnicity is identified, there is no race to accompany it.**  
    - We cannot explicitly analyze the influence of ethnicity on a specific race, but rather on all races.  
    - For this reason, we will create two separate variables relating to race/ethnicity:  
        - One that includes race and ethnicity together.  
        - Another that specifies Hispanic or not without any racial measure.

2. **Additionally, we only have information about sex, since these data are from death certificates.**  
    - As such, we cannot consider the influence of other potentially relevant factors like one's gender identity.

3. **We cannot examine how sexual orientation impacts opioid mortality.**  
    - There is no measure of this either.

_It is important to note that these data are taken from death certificates, but do not include any potential identifying data._ 



#### Subsetting the Data

The data was organized in a way that required a considerable amount of cleaning and subsetting in order to isolate the measures we intended to explore. Specifically, the demographic measures do not have their own variables, but rather are compiled into one variable, called **"STUB_LABEL"**. 

We must create another variable called **"STUB_NAME"** and the corresponding numerical variables **"STUB_LABEL_NUM"** and **"STUB_NAME_NUM"** to relabel the required variables and subset them accordingly for our analysis. 

    - Each "STUB_NAME_NUM" number corresponds to categorical measure(s), for the sake of our analysis, we will focus on the "Sex and race and Hispanic origin" level of the "STUB_NAME" variable which corresponds to 5 in "STUB_NAME_NUM."
    
    -A similar categorization system exists for specific groups with the "STUB_LABEL" and "STUB_LABEL_NUM" corresponding to "Female: Not Hispanic or Latino: White" and 5.7, respectively. 
    
    -The "STUB_LABEL_NUM" builds upon the "STUB_NAME_NUM" with a decimal number to denote a specific sex, ethnicity, and race. 


***For this study we subset the data to exclusively include observations with specified sex, race, and ethnicity.***

## Methods

#### Overview 

In order to evaluate how to best analyze the data we must familiarize ourselves with the variables and how they're oriented.

This data set classifies drug overdose by type, and also takes into account estimations for the rate of overdose per 100,000 residents for group listed.


1. **Relabel the type of drug overdose for clarity**  
   This relies upon the existing **“PANEL”** and **“PANEL_NUM”** variables.  
   - We will produce a table with these two variables so we can see which number correlates with a specific drug type.  

2. **Subset the data to focus on the influence of race and sex on drug overdose**  
   - Race and gender are under **“STUB_LABEL_NUM”**, so we will disaggregate it and rename it to make the data more readable and optimized for our statistical tests.  

3. **End with five variables of interest**  
   - **“ESTIMATE”**: Provides the average rate of deaths per 100,000 people.  
   - **“Race”** and **“Sex”**: Show the racial identity and sex classification provided on the death certificate.  
   - **“YEAR”**: Notes the year in which the deaths were registered.  
   - **“Responsible_Drug”**: Identifies the drug responsible for the death as noted on the death certificate.  


#### Loading in the Data and Packages

In order to answer our research question we need to **load the data and relevant packages** like tidyverse, ggplot2 and gridExtra that will allow us to complete the previously mentioned subsetting and the statistical tests required to compare the estimated death rates between certain groups.

```{r setup}
# Packages used 
library(tidyverse)
library(ggplot2)
library(gridExtra)

# Loading the dataset
drug <- read_csv("Renv/Data/drug_data.csv", show_col_types = F)

```


#### Examining and Cleaning the Data

Identifying how each drug overdoes type is categorized so we can relabel it. 

```{r}
table(drug$PANEL_NUM, drug$PANEL)
```

**Number** | **Description**  
--- | ---  
0 | all drug overdose deaths regardless of drug type 
1 | drug overdose deaths from any opioids  
2 | any drug overdose death from synthetic opioids    
3 | drug overdose deaths from methadone     
4 | drug overdose deaths from synthetic opioids except methadone   
5 | drug overdose deaths from heroin  


```{r}
# Filtering to see the impact of the death rate estimate variation by racial/gender groups based on type of drug causing overdose

drug_type <- drug %>% 
  rename(Responsible_Drug = "PANEL_NUM") %>% 
  mutate(
     Responsible_Drug = case_when(
      Responsible_Drug == 0 ~ "All Drug Types",
      Responsible_Drug == 1 ~ "Any Opioid",
      Responsible_Drug == 2 ~ "Natural/Semisynthetic Opioid",
      Responsible_Drug == 3 ~ "Methadone",
      Responsible_Drug == 4 ~ "Synthetic Opioid",
      Responsible_Drug == 5 ~ "Heroin",
      TRUE ~ "Unknown" 
     ))

# Factor function to maintain the order of drug levels in visualization

drug_type$Responsible_Drug <- factor(drug_type$Responsible_Drug, levels =c("All Drug Types", "Any Opioid", "Natural/Semisynthetic Opioid", "Methadone", "Synthetic Opioid", "Heroin"))

```

**Production of tables to clarify the required demographic data, looking at the number that is populated by observations.**

**Next we will subset the data from the larger data set that are organized by race and sex**

```{r}
# Finding with single digit number represents the racial/sex level of the STUB_NAME variable 

table(drug$STUB_NAME_NUM, drug$STUB_NAME)

# 4 is the number that corresponds to Race and Sex, so we filtered on this to see which decimal number corresponds to each racial/gender group. 
drug_exam <- drug_type %>% 
  filter(STUB_NAME_NUM == 4)

# Produce a table to see which number corresponds to a specific racial/sex pairing (i.e Black Male)
table(drug_exam$STUB_LABEL_NUM, drug_exam$STUB_LABEL)

```

**Number** | **Description**  
--- | --- 
4.1 | White Males
4.2 | Black Males 
4.3 | Native American Males  
4.4 | Asian Males     
4.5 | White Females   
4.6 | Black Females  
4.7 | Native American Females
4.8 | Asian Females

This allows for the **creation of a new variable** that explicitly states that states these race/sex pairings and produce a subset that focus on this new variable.

The **subset disaggregates race and sex** so that we can perform a ***2-way ANOVA*** to see if these two variables are statistically _significant predictors for drug overdose mortality_. 


```{r}
drug_RES <- drug_type %>% 
  filter(STUB_NAME_NUM == 5) %>%
  rename(Race_Ethnicity = "STUB_LABEL_NUM") %>% 
  mutate(
     Race_Ethnicity = case_when(
     Race_Ethnicity == 5.1 ~ "Hispanic, Unspecified Race",
      Race_Ethnicity == 5.2 ~ "NH White",
      Race_Ethnicity == 5.3 ~ "NH Black",
      Race_Ethnicity == 5.4 ~ "NH Native American",
      Race_Ethnicity == 5.5 ~ "NH AAPI",
      Race_Ethnicity == 5.6 ~ "Hispanic, Unspecified Race",
      Race_Ethnicity == 5.7 ~ "NH White",     
      Race_Ethnicity == 5.8 ~ "NH Black",
      Race_Ethnicity == 5.9 ~ "NH Native American",
      Race_Ethnicity == 5.91 ~ "NH AAPI",
      TRUE ~ "Unknown" 
    )) %>% 
   mutate(
   Ethnicity = case_when(
      grepl("Black", Race_Ethnicity) ~ "Non-Hispanic",
      grepl("White", Race_Ethnicity) ~ "Non-Hispanic",
      grepl("Native American", Race_Ethnicity) ~ "Non-Hispanic",
      grepl("AAPI", Race_Ethnicity) ~ "Non-Hispanic",
      grepl("Hispanic", Race_Ethnicity) ~ "Hispanic, Unspecified Race",
      TRUE ~ "Unknown"
    )) %>% 
  mutate(
    Sex = case_when(
      grepl("Female", STUB_LABEL) ~ "Female",
      grepl("Male", STUB_LABEL) ~ "Male",
      TRUE ~ "Unknown")) 

##Further cleaning data to remove NAs, Unknowns, and generalized drug types
drug_RES$FLAG <- NULL
drug_RES <- drug_RES %>% 
  filter(!grepl("Unknown", Race_Ethnicity)) %>% 
  filter(!grepl("Any", Responsible_Drug)) %>% 
  filter(!grepl("All", Responsible_Drug)) %>% 
  na.omit


##Disaggregating by Sex for Central Measures and Visualization
drug_Fem <- drug_RES %>% 
  filter(Sex == "Female")

drug_Male <- drug_RES %>% 
  filter(Sex == "Male")
```


## Visualizing the Distribution

#### All Racial and Sex Identity Groups, for all Drug Overdose Types, from 1999 - 2018

 We produced this visualization initially to get an **idea of the increase of drug mortality over around 20 years**. This time period is especially relevant as the CDC declares that this time range represents the "[Opioid Epidemic](https://www.cdc.gov/museum/pdf/cdcm-pha-stem-uncovering-the-opioid-epidemic-lesson.pdf)"and many of the drugs identified in this data set are classified as some form of an opioid. This information is also helpful in highlighting the variation in mortality by race/sex pairings.
 

##### ***Plots Comparing Men and Women Across Race/Ethnic Groups***

```{r, echo = F}

overall_plot_M <- ggplot(drug_Male, aes(x = factor(YEAR), y = ESTIMATE, fill = Race_Ethnicity)) +
  geom_bar(stat = "identity", position = "dodge", na.rm = TRUE) +
  labs(title = "Estimates for Males by Identity and Year",
       x = "Year",
       y = "Estimated Deaths per 100,000") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90)) + labs(fill = "Race/Ethnic Identity") +  theme(plot.title = element_text(hjust = 0.5)) + ylim(0,20)


overall_plot_F <- ggplot(drug_Fem, aes(x = factor(YEAR), y = ESTIMATE, fill = Race_Ethnicity)) +
  geom_bar(stat = "identity", position = "dodge", na.rm = TRUE) +
  labs(title = "Estimates for Females by Identity and Year",
       x = "Year",
       y = "Estimated Deaths per 100,000") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90)) + labs(fill = "Race/Ethnic Identity") +  theme(plot.title = element_text(hjust = 0.5))+ ylim(0,20)

grid.arrange(overall_plot_F, overall_plot_M)
```


**Conclusion:** The distribution of the data shows that compared to females, ***males experience almost an exponential increase in deaths due to drug overdose spiking around the late 2000s.***

**Importance:** These data tell us a lot regarding the different observations witnessed. The distribution provide us with essential information about certain sex/race pairings, as well as the general trends of these groups over time. Such information is helpful as we perform more "narrow" visualizations and analysis. 


##### ***Distribution of Sex By Estimate ***

```{r, echo = F}
year_Sex <- ggplot(drug_RES, aes(x = factor(YEAR), y = ESTIMATE, fill = Sex)) +
  geom_boxplot(na.rm = TRUE) +
  labs(title = "Distribution of Overdose Estimates by Sex and Year", 
       x = "Year", 
       y = "Estimated Deaths per 100,000") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90)) + theme(plot.title = element_text(hjust = 0.5))

year_Sex
```

**Conclusion:** The graph shows that a common trend for both female and male is that ***as the year increased the number of drug overdose increased.*** While it was a subtle increase for females, the males saw a larger rise in drug overdose deaths. 

The ***differences in the rate of increase are visible from the continued distance in the median bar between females and males as time goes on***. We can verify this with additional central tendency measures. 


##### ***Distribution of Overdose Estimates by Drug type and Race/Ethnic Identity ***

We wanted to see the **_distribution of the different drug type overdose among the different race and genders to see which type is more prevalent_**.

```{r, echo= FALSE}
type_RE <- ggplot(drug_RES, aes(x = factor(Responsible_Drug), y = ESTIMATE, fill = Race_Ethnicity)) +
  geom_bar(stat = "identity", position = "dodge",na.rm = TRUE) +
  labs(title = "Overdose Estimates by Drug Type and Race/Ethnic Identity", 
       x = "Drug Type", 
       y = "Estimated Deaths per 100,000") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90)) + labs(fill = "Race/Ethnic Identity") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0,20)

type_RE
```

**Conclusion:** The graph shows that among the drug types, **synthetic opioids appear to be one of the leading causes of death** by drug overdose. Specifically for Non-Hispanic White and Non-Hispanic black as they have a high amount of death compared to other Race/Ethnic identities.

We can also see that **methadone has the lowest rates for every group** when averaging rates from 1999-2018.This is likely because methadone is typically **given out as clinics as way to mitigate withdrawal symptoms** for drug users trying to **decrease or eliminate their opioid consumption**. Since clinics typically give this drug out, it is in a more controlled environment with a measure dose which is why the rates are lower


##### ***Distribution of Overdose Estimates by Drug type and Race/Ethnic Identity Among Men and Women***

```{r, echo=FALSE}
type_REF <- ggplot(drug_Fem, aes(x = factor(Responsible_Drug), y = ESTIMATE, fill = Race_Ethnicity)) +
  geom_bar(stat = "identity", position = "dodge",na.rm = TRUE) +
  labs(title = "Overdose Estimates by Drug Type and Race/Ethnic Identity (Women)", 
       x = "Drug Type", 
       y = "Estimated Deaths per 100,000") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90)) + labs(fill = "Race/Ethnic Identity") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0, 20)
type_REF
```

```{r, echo=FALSE}
type_REM <- ggplot(drug_Male, aes(x = factor(Responsible_Drug), y = ESTIMATE, fill = Race_Ethnicity))+
  geom_bar(stat = "identity", position = "dodge",na.rm = TRUE) +
  labs(title = "Overdose Estimates by Drug Type and Race/Ethnic Identity (Men)", 
       x = "Drug Type", 
       y = "Estimated Deaths per 100,000") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90)) + labs(fill = "Race/Ethnic Identity") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0, 20)
type_REM
```


**Conclusion**: The first two categories have the lowest estimates likely because they are more inclusive/broader categories in general. However, the **synthetic opioid is the highest out of the rest of the categories**. Fentanyl is a highly potent and potentially lethal drug that is classified as a synthetic opioid. Although we cannot conclude based upon the language used in this data set, **fentanyl is likely the reason why this categories’ estimates are higher**. 

Another important note is that this is the only category where **Black males have a higher estimate than White men**, which may indicate a _need for further inquiry_. Among the rest of the drug type categories, the same trends by race/sex pairing are generally seen.



## Viewing Central Tendency of the Outcome Variable 

```{r}
## To understand the information from the data, we have to understand the outcome variable which showcase the estimate for drug overdose death, these aggregrate measures are for all race/sex pairings across time, and provided a "standard" of sorts when comparing mean, median, or standard deviation for a specific year, race, and/or sex group. 


mean_estimate <- mean(drug_RES$ESTIMATE, na.rm = TRUE) #4.2
median_estimate <- median(drug_RES$ESTIMATE, na.rm = TRUE) #2
std_estimate <- sd(drug_RES$ESTIMATE, na.rm = TRUE) #5.19




## To gain a better understanding of the outcome, we chose to focus on the mean of the different races and gender to see the mean of drug overdose death rate amongst the different groups as well as the standard deviation 

meanrace <- drug_RES %>%
  group_by(Race_Ethnicity, Sex) %>%
  summarize(mean_estimate = mean(ESTIMATE, na.rm = TRUE), .groups = 'drop')

meanrace

stdrace<- drug_RES %>%
  group_by(Race_Ethnicity, Sex) %>%
  summarize(std_dev_estimate = sd(ESTIMATE, na.rm = TRUE), .groups = 'drop')

stdrace

##Racial Group Mean and Standard Deviation by Year

mean_year_race <- drug_RES %>% 
  group_by(YEAR, Race_Ethnicity) %>% 
   summarize(mean_estimate = mean(ESTIMATE, na.rm = TRUE), .groups = 'drop')

mean_year_race

std_year_race <- drug_RES %>% 
  group_by(YEAR, Race_Ethnicity) %>% 
    summarize(std_dev_estimate = sd(ESTIMATE, na.rm = TRUE), .groups = 'drop')

std_year_race

#Sex Group Mean and Standard Deviation by Year

mean_year_sex <- drug_RES %>% 
  group_by(YEAR, Sex) %>% 
   summarize(mean_estimate = mean(ESTIMATE, na.rm = TRUE), .groups = 'drop')

mean_year_sex

std_year_sex <- drug_RES %>% 
  group_by(YEAR, Sex) %>% 
     summarize(std_dev_estimate = sd(ESTIMATE, na.rm = TRUE), .groups = 'drop')

std_year_sex

##Drug Type Mean and Standard Deviation by Year 
mean_year_type <- drug_RES %>% 
  group_by(YEAR, Responsible_Drug) %>% 
  filter(!Responsible_Drug %in% c("All Drug Types", "Any Opioid")) %>% 
  summarize(mean_estimate = mean(ESTIMATE, na.rm = TRUE), .groups = 'drop')

mean_year_type

std_year_type <- drug_RES %>% 
  group_by(YEAR, Sex) %>% 
  filter(!Responsible_Drug %in% c("All Drug Types", "Any Opioid")) %>% 
  summarize(std_dev_estimate = sd(ESTIMATE, na.rm = TRUE), .groups = 'drop')

std_year_type



```

#### The central tendencies produced in this section give us a numerical representation of some of the trends seen in the previous descriptive visualizations. Like the graphs showed, White Males and Native American Males have the highest mean rates across drug type and year with an estimate of 6.98 and 6.05 deaths per 100,000 people on average, respectively. White Males estimated death rate is nearly ten times that of Asian women, showing the range possible based on racial identity and sex. Like previously mentioned, White Males and Black Males have the highest standard deviation of estimates, with 7.33 and 6.71, respectively. Across racial groups, we see that females consistently have lower average estimated death rates and standard deviations which point to the influence of sex. We see a fairly consistent increase in mean and standard deviation as the year increases, although the increase "steepends" as the years approach 2018. These increases were consistent regardless of sex and race. Consistent increases were seen when focusing on drug type as well. We excluded the two more general categories of "All Drug Type" and "Any Opioid" here to focuse on more narrowly defined drug type categories. During the earlier years, methadone and synthetic opioids "trade" places with which level has the lowest estimate. This is until 2014, when sythetic opioid deaths become much more prevalent. This is likely due to the increase in fentanyl-related drug deaths during the same time. Despite synthetic opioids being having some of the lowest rates from 1999-2013, it is the highest of the narrowly defined levels due to the sharp increase in deaths from 2014-2018. 


## ANOVA Test

#### Given that we have a numerical outcome variable and multiple categorical variables, we decided that ANOVA would be the ideal stastical test to run on our data. 

### Initial Test
```{r, echo = T}
## A Two Way ANOVA with Race and Sex as the categorical variables of analysis
RS_Model <- aov(ESTIMATE ~ Race_Ethnicity + Sex,
  data = drug_RES
)
summary(RS_Model)

table(drug_RES$Sex, drug_RES$Race_Ethnicity)

m1 <- aov(ESTIMATE ~ Ethnicity, data = drug_RES)
m2 <- aov(ESTIMATE ~ Ethnicity + Race_Ethnicity, data = drug_RES)
m3 <- aov(ESTIMATE ~ Race_Ethnicity + Sex, data = drug_RES)
m4 <- aov(ESTIMATE ~ Ethnicity + Race_Ethnicity + Sex, data = drug_RES)
m4.1 <- aov(ESTIMATE ~ Ethnicity + Race_Ethnicity + Sex + Sex:Race_Ethnicity + Responsible_Drug, data = drug_RES)
m5 <- aov(ESTIMATE ~ Ethnicity + Race_Ethnicity + Sex + Responsible_Drug, data = drug_RES)
m6 <- aov(ESTIMATE ~ Ethnicity + Race_Ethnicity + Ethnicity:Race_Ethnicity, data = drug_RES)
m7 <- aov(ESTIMATE ~ Ethnicity + Race_Ethnicity + Race_Ethnicity:Responsible_Drug, data = drug_RES)
m8 <- aov(ESTIMATE ~ Ethnicity + Race_Ethnicity + Responsible_Drug + Race_Ethnicity:Responsible_Drug, data = drug_RES)

summary(m1)
summary(m2)
summary(m3)
summary(m4.1)
summary(m4)
summary(m5)
summary(m6)
summary(m7)
summary(m8)

## From the table we can see this is somewhat of a balanced Two-way ANOVA as we can see close to the same number of observations across the different categories
```

#### A two-way ANOVA conducted between Race and Sex in relation to the outcome of drug overdose. The H0 (null hypothesis) suggest that there is no difference of the means between the 2 groups : u1=u2=u3. While the Ha (alternative hypothesis) suggest that there is a difference between at least one of the means of the 2 groups. The results indicate that both Race and Sex are significant to the outcome of drug overdose. With a high F-value and small p-value below the alpha level of 0.5, both Race and Sex are significant effects of the outcome.

#### However the amount of residuals in our test suggest there is still another factor that explains the variability in the outcome of drug overdose.

### Inclusion of Iteraction Term

```{r, echo = T}
## A Two Way ANOVA with Race and Sex as the categorical variables of analysis, with an additional consideration for an interaction effect between these two variables
RS_IntModel <- aov(ESTIMATE ~ Race_Ethnicity + Sex + Race_Ethnicity:Sex,
  data = drug_RES
)

summary(RS_IntModel)
```
#### Based on the result, we can see that the interaction didn't provided any statistical significance to the outcome of the drug overdose. This is indicated from the low f-value an da high p-value over 0.05. The interaction term with a p-value of 0.065 indicates that we fail to reject the null hypothesis.

### Inclusion of an Additional Predictor 

```{r, echo = T}
## A Three Way Anova with Race, Sex, and Responsible Drug as the categorical variable of analysis 
Type_Model<- aov(ESTIMATE ~ Race_Ethnicity + Sex + Responsible_Drug,
  data = drug_RES)
summary(Type_Model)




```

#### Lastly, we conducted a three way ANOVA we wanted to determine whether or not the Responsible_drug factor played a role in the outcome of drug overdoses. Based on the results we can see that inclusion of responsible drug is statistically significant. The f-values of 161.09 and a p-value of <2e-16, showcased that the Responsible_drug, or the type of drug, plays a role in drug overdose deaths. We can also see a decrease in the residuals indicating, that the Responsible_Drug variable accounted for some of the variability of the Estimate outcome variable.  



#### Drug overdoses in the US have been steadily increasing over the years. However the distribution of these deaths are not uniform for everyone. Certain groups, whether they be based around age, race, or sex, can be more prone than others to have a drug over dose. To better understand which groups are most at risk and for what drugs they are most likely to be impacted by we produced multiple descriptive visualizations and ANOVA tests on data taken from the National Center for Health Staitstics and specifically their National Vital Statistics System. These Data were gathered in 2019 and are based off of information taken from death certificates. The official information provided about the data set on the Center for Disease Control's website is, "Drug overdose death rates, by drug type, sex, age, race, and Hispanic origin: United States, selected years 1999-2018". Additional information about these specific data can be found here: https://data.cdc.gov/NCHS/Drug-overdose-death-rates-by-drug-type-sex-age-rac/95ax-ymtc/about_data. The CSV formatted data can be found here: https://www.cdc.gov/nchs/hus/contents2019.htm#Table-008. Chart visualizations of the data are available here: https://www.cdc.gov/nchs/data/hus/2019/008-508.pdf. It is important to note that there is no gender variable in this data set, but rather a sex variable. As such, we cannot confirm or deny if gender minorities, like non-binary individiuals, transgender men, and transgender women are included in the data set, which reduces the inclusion and representation of these data. To avoid confusion regarding this, we will explicitly refer to this variable as sex or sexual identity as to not misrepresent the data. For the sake of the present analysis, we focus on race and sexual identity, this is because of the intersectional nature of these two identities and how they are tied to systemic disaprities. Moreover, the selection of these groups will enable more targeted interventions based upon the findings.  
